Something about Continuous Deductive Reasoning over Streamed RDF Data, maybe claim Soundness or some other Theoretical Quality

### 2015-07-15 content/contribution/continuous-datalog.tex
removed rogue FIXME (actual final change to section 2)
Some final alterations to bring the entire paper more fluidly in line, terminologically, and to balance the placement of figures more evenly through the document.

\begin{nestedsection}{Continuous Datalog}{semantics}
The first requirement of stream reasoning is a well-defined semantics
that takes into account both the continuous nature of a subset of the
base axioms of a stream reasoning problem, as well as its
instantaneous truth and entailments, in line with the work of the
reasoning community.  However, it must also provide a continuous
interpretation of those instantaneous entailments in order to
integrate with other stream-processing technologies.  More
specifically, an appropriate semantics for stream reasoning on the
Semantic Web must: allow the expression of OWL 2 ontologies to some
degree of expressivity; not sacrifice the precise semantics thereof
whilst incorporating the semantics of stream processing; and allow for
the continuous derivation of entailments in a concise manner.

Continuous Datalog (henceforth C-Datalog) is an extension of positive
Datalog with stream-processing semantics, with a view to accommodating
the aims of the latter within the well-defined model-theoretic
semantics of the former, which are shared with RIF-Core
\cite{w3crifbld} and therefore support reasoning at the expressivity
of OWL 2 RL. As part of its application to Semantic Web reasoning, we
assume that the atomic formulae of the Datalog programs described
below correspond to RDF triples: $\langle subj, pred, obj\rangle$.

C-Datalog differs from conventional Datalog through the addition of
sources of continuously generated or reported facts (i.e streams), and
the application of sliding window semantics over those sources,
thereby providing for the interpretation of C-Datalog programs as
sequences of instantaneous states that are each valid positive Datalog
programs.  As such, these states each inherit the semantics of
positive Datalog.  Furthermore, it involves the continuous derivation
of entailments from this sequence of instantaneous states as a stream
of facts; these streams share a data model with the streams of
continuous queries, as defined by the stream processing community,
thereby allowing the integration of C-Datalog programs into stream
processing pipelines. Finally, we show that these streams may be
manipulated to retrieve the instantaneous sets of entailments produced
by the equivalent Datalog programs.

@@

\begin{axiom}\label{axiom:continuous datalog: window range leq 0}
A window of range $0$ over a stream $S$ at an instant 
${i \in \mathbb{T}}$ is always empty.
\begin{equation*}
W^{S,0}_{i} = \varnothing
\end{equation*}
\end{axiom}

\begin{axiom}\label{axiom:continuous datalog: window composition}
A window of range ${x > 0}$ over a stream $S$ at an instant 
${i \in \mathbb{T}}$ is equal to the union of the $x$ consecutive
windows of range $1$ over the stream $S$ preceding and including that at
instant $i$.
\begin{equation*}
W^{S,x}_{i} = \mathop{\cup}_{j=0}^{x-1} W^{S,1}_{i-j}
\end{equation*}
\end{axiom}

\begin{axiom}\label{axiom:continuous datalog: window disjointness}
A window of range ${x \in \mathbb{N}}$ over a stream $S$ at an
instant ${i \in \mathbb{T}}$ is disjoint from a second window of 
range ${y \in \mathbb{N}}$ over $S$ at any instant before $i - x$.
\begin{equation*}
W^{S,x}_{i} \cap W^{S,y}_{i-z} = \varnothing
\end{equation*}
where $z \geq x$.
\end{axiom}

@@

We consider the entailment of the WC-Datalog program $\Pi^{C,r}$ not
as a single stream of fact instances (as for C-Datalog), but in three
parts: a set of fact instances $E^0 = cons^T(\langle R, F^0\rangle)$
that is the entailment of the TA-Datalog program $\langle R,
F^0\rangle$; a stream $P$ of fact instances entailed by the rules $R$ and
the fact instances in $F^0$ and $S$, ordered by the instant at which they
became justified (i.e. their instants); and a stream $N$ of those same fact
instances, ordered by the instant at which they ceased to be justified:
$\langle R, F^0, S, r \rangle \models \langle E^0, P, N \rangle$. $P$
and $N$ can be considered to be the continuous reasoning equivalents of
the streams produced by the \emph{IStream} and\emph{DStream} operators
respectively of the \emph{CQL} continuous query language \citep{CQL},
which create streams based on the additions to and deletions from a
relation. Note that $P$ is an in-order stream whose members are a subset
of those in $E^S$ for an equivalent C-Datalog program, while $N$ may be
an out-of-order stream.

%%%FIXME
In order to

@@

\begin{axiom}\label{axiom:continuous datalog: entailment disjointness}

If two sliding windows over the same in-order stream $S$ cover
non-overlapping time periods, then the set of fact instances entailed
by ${\langle R, W^{S,x}_i \rangle}$ is necessarily disjoint from the
set of fact instances entailed by $\langle R, W^{S,y}_{i-z} \rangle$,
where ${z \geq x}$.
\[
\{ e \mid \langle R , W^{S,x}_i \rangle \models e \} \cap \{ e \mid \langle R , W^{S,y}_{i-z} \rangle \models e \} = \varnothing
\]
\end{axiom}

@@

The instantaneous entailment of a $\Pi^{C,r}_i$ may be divided into two
semantically distinct portions: the \emph{persistent} portion is the
entailment of ${\langle R, F^0 \rangle \models E^0}$, as in a WC-Datalog
program, which is guaranteed to be part of the entailment of every state
due to the monotonicity of positive Datalog; the \emph{transitive} portion
is the set $E^T_{i}$ of entailed fact instances whose justifications are at
least partially contained in $W^{S,r}_i$. We define $E^T_i$ as:
\[
E^T_i = \{ x \mid \Pi^{C,r}_i \models x \land J^T \cup W^{S,r}_i \neq \varnothing \}
\]

\noindent where $J^T$ is a justification for the entailment of x by
$\Pi^{C,r}_i$. By this definition and Axiom~\ref{axiom:continuous
  datalog: entailment disjointness}, $E^0$ is guaranteed to be
disjoint from every $E^T_i$ for every state $\Pi^{C,r}_i$ of every
WC-Datalog program $\Pi^{C,r}$.
\end{definition}

@@

\begin{axiom}
\label{axiom:continuous datalog: positive window increment}
The window of range 1 at any given instant ${i \in \mathbb{T}}$ over
the stream $P$ entailed by a WC-Datalog program $\Pi^{C,r}$ is defined
as the set of fact instances that are entailed by the state
$\Pi^{C,r}_i$ that were not entailed by the state $\Pi^{C,r}_{i-1}$.
\begin{equation*}
W^{P,1}_{i} = \left( E^0 \cup E^T_{i} \right) \setminus
\left( E^0 \cup E^T_{i-1} \right) = E^T_{i} \setminus E^T_{i-1}
\end{equation*}
\end{axiom}

\begin{axiom}
\label{axiom:continuous datalog: negative window increment}
The window of range 1 at any given instant ${i \in \mathbb{T}}$ over
the stream $N$ entailed by a WC-Datalog program $\Pi^{C,r}$ is defined
as the set of fact instances that were entailed by the state
$\Pi^{C,r}_{i-1}$ that are no longer entailed by the state
$\Pi^{C,r}_i$.
\begin{equation*}
W^{N,1}_{i} = \left( E^0 \cup E^T_{i-1} \right) \setminus
\left( E^0 \cup E^T_{i} \right) = E^T_{i-1} \setminus E^T_{i}
\end{equation*}
\end{axiom}

\begin{axiom}
\label{axiom:continuous datalog: entailment precedes negation}
Fact instances may only be negated by some part of their justification
leaving the sliding window $W^{S,r}_{i}$, this necessarily being after
they were entailed.  As such, a window of any range ${x \in
  \mathbb{N}}$ over the stream of entailments $P$ at any given instant
${i \in \mathbb{T}}$ is disjoint from a window of any range ${y \in
  \mathbb{N}}$ over the stream of entailment negations $N$ at any
instant before ${i - x}$.
\begin{equation*}
  W^{P,x}_{i} \cap W^{N,y}_{i-z} = \varnothing
\end{equation*}
where $z \geq x$.
\end{axiom}

@@

However, it follows from these axioms that the instantaneous set
$E^T_{i}$ of instances entailed at any given instant $i$ by any
$\Pi^{C,r} = \langle R, F^0, S, r \rangle$ may be reconstructed from
windows of sufficient range over its entailed streams $P$ and $N$ at
$i$:
\[ E^T_{i} = W^{P,x}_{i} \setminus W^{N,y}_{i} \]
where ${y \geq x \geq r}$. The minimum range $x$ is such to achieve
the \emph{complete} set of entailment instances $E^T_{i}$ of the
$\Pi^{C,r}_{i}$. In addition, the minimum range $y$ is such to achieve the
\emph{sound} set $E^T_{i}$. As such, a \emph{valid window} that
contains only and exactly those entailed instances $E^T_{i}$ at every
instant $i$ may be trivially maintained from the streams $P$ and $N$
continuously entailed by some $\Pi^{C,r}$.

% \begin{thrm}\labelhyp{continuous datalog: entailment} \hfill \\

% For any ${\Pi^{C,r} \models \langle P, N \rangle}$, where ${\Pi^{C,r}_{i} \models E^0 \cup E^T_{i}}$ for any ${i \in \mathbb{T}}$, the set of facts in a window of any range ${x \geq r}$ over the stream $P$ at a given instant ${i}$ and not in a window of any range ${y \geq x}$ over the stream $N$ at the same instant $i$ is equal to the transitive portion $E^T_{i}$ of the entailment of the state of the ${\Pi^{C,r}}$ at that instant $i$:

% \begin{equation*}
%   W^{P,x}_{i} \setminus W^{N,y}_{i} = E^T_{i}
% \end{equation*}

% \end{thrm}

% \begin{proof}

% In order to prove \refhyp{continuous datalog: entailment} for the specific case of ${x \geq r}$, for some range $r$ of the window over $S$ in some ${\Pi^{C,r} = \langle R,F^0,S,r \rangle}$, a general case must first be proven for all ${x \geq 1}$.
% This general case constitutes the Lemma \ref{hyp:continuous datalog: entailment lemma}, which provides an equivalence between the entailments of discrete states of $\Pi^{C,r}$ and windows over the streamed entailment ${\langle P, N \rangle}$ for all window ranges ${x \geq 1}$.

% \begin{lem}\label{hyp:continuous datalog: entailment lemma} \hfill \\

% For any ${\Pi^{C,r} \models \langle P,N \rangle}$, where ${\Pi^{C,r}_{i} \models E^0 \cup E^T_i}$ for any ${i \in \mathbb{T}}$, the set of facts in a window of any range ${x \geq 1}$ over the stream $P$ at a given instant ${i}$ and not in a window of any range ${y \geq x}$ over the stream $N$ at the same instant $i$ is equal to the transitive portion $E^T_i$ of the entailment of the state of $\Pi^{C,r}$ at that instant $i$ less the intersection of the transitive portions of the entailments of all previous states of $\Pi^{C,r}$, from instant ${i-1}$ up to and including ${i-x}$:

% \begin{equation*}
%   W^{P,x}_{i} \setminus W^{N,y}_{i} = E^T_{i} \setminus \mathop{\cap}^{x}_{d=1} E^T_{i-d}
% \end{equation*}

% \end{lem}

% The inductive proof of Lemma \ref{hyp:continuous datalog: entailment lemma} is available in \ref{app:semantics: proof for entailment}:
% a \emph{general case} (\reffig{semantics: induction general case}) eliminates the range ${y \geq x}$ over $N$ as a factor;
% a \emph{base case} (\reffig{semantics: induction base case}) is shown to hold when ${x = 1}$;
% and the \emph{inductive case} (\reffig{semantics: induction inductive case}) holds when ${x = k + 1}$, assuming the lemma holds for ${x = k}$.
% Given this lemma, the derivation of \refhyp{continuous datalog: entailment} presented in \reffig{continuous datalog: proof for entailment} follows.

% \begin{figure*}

% \centering

% \begin{align*}
%   E^T_{i} & = W^{P,x}_{i} \setminus W^{N,y}_{i} \\
%   & = E^T_{i} \setminus \mathop{\cap}^{x}_{d=1} E^T_{i-d}
%   && \text{, Lemma~\ref{hyp:continuous datalog: entailment lemma}} \\
%   & = \mathop{\cup}^{x}_{d=1} \left( E^T_{i} \setminus E^T_{i-d} \right) \\
%   & = \left( E^T_{i} \setminus E^T_{i-x} \right) \cup
%   \mathop{\cup}^{x-1}_{d=1} \left( E^T_{i} \setminus E^T_{i-d} \right) \\
%   & = E^T_{i} \cup \mathop{\cup}^{x-1}_{d=1} \left( E^T_{i} \setminus E^T_{i-d} \right)
%   && \text{, Axiom~\ref{axiom:continuous datalog: entailment disjointness}} \\
%   & = E^T_{i} \cup \left( E^T_{i} \setminus \mathop{\cap}^{x-1}_{d=1} E^T_{i-d} \right) \\
%   & = E^T_{i} && \qedhere
% \end{align*}

% \caption{Derivation for proof of \refhyp{continuous datalog: entailment}}
% \labelfig{continuous datalog: proof for entailment}

% \end{figure*}

% \end{proof}

@@

\begin{figure*}
\centering
$
\begin{CD}
	{\Pi^C = \langle R, F, S \rangle} @>\models>> {E^S} \\
	@V{add\ window}VV \\
        {\Pi^{C,r} = \langle R, F^0, S, r \rangle} @>\models>> {\langle cons^T(\langle R, F^0\rangle), P, N \rangle} \\
        @V{evaluate\ window}VV @VV{evaluate\ window}V \\
        {\Pi^{C,r}_i = \langle R, F^0 \cup W^{S,r}_i \rangle} @>\models>> {cons^T(\langle R, F^0 \rangle) \cup (W^{P,r}_i \setminus W^{N,r}_i)} \\
	@VVV @VVV \\
        {\Pi^T = \langle R, F^T \rangle } @>\models>> {cons^T(\Pi^T)} \\
	@V{strip}VV @VV{strip}V \\
        {\Pi = \langle R, F \rangle } @>\models>> {cons(\Pi)}
\end{CD}
$
\caption{Commutative diagram from Continuous Datalog programs to Datalog entailments.}
\labelfig{CD-TD-D}
\end{figure*}
\end{nestedsection}

>>>>

\begin{nestedsection}{Semantics: Windowed Continuous Datalog}{semantics}
The first requirement of stream reasoning is a well-defined semantics
that takes into account both the continuous nature of a subset of the
base axioms of a stream reasoning problem, as well as its
instantaneous truth and entailments, in line with the work of the
reasoning community.  However, it must also provide a continuous
interpretation of those instantaneous entailments in order to
integrate with other stream-processing technologies.  More
specifically, an appropriate semantics for stream reasoning on the
Semantic Web must: allow the expression of OWL 2 ontologies to some
degree of expressivity; not sacrifice the precise semantics thereof
whilst incorporating the semantics of stream processing; and allow for
the continuous derivation of entailments in a concise manner.

Windowed Continuous Datalog (henceforth WC-Datalog) is an extension of positive
Datalog with stream-processing semantics, with a view to accommodating
the aims of the latter within the well-defined model-theoretic
semantics of the former, which are shared with RIF-Core
\cite{w3crifbld} and therefore support reasoning at the expressivity
of OWL 2 RL. As part of its application to Semantic Web reasoning, we
assume that the atomic formulae of the Datalog programs described
below correspond to RDF triples: $\langle subj, pred, obj\rangle$.

WC-Datalog differs from conventional Datalog through the addition of
sources of continuously generated or reported facts (i.e streams), and
the application of windowing semantics over those sources,
thereby providing for the interpretation of WC-Datalog programs as
sequences of instantaneous states that may each be reduced to valid positive Datalog
programs.  As such, these states each inherit the semantics of
positive Datalog.  Furthermore, it involves the continuous derivation
of entailments from this sequence of instantaneous states as a stream
of facts; these streams share a data model with the streams of
continuous queries, as defined by the stream processing community,
thereby allowing the integration of WC-Datalog programs into stream
processing pipelines. Finally, we show that these streams may be
manipulated to retrieve the instantaneous sets of entailments produced
by the equivalent Datalog programs.

@@

{\nobreak\begin{axiom}\label{axiom:continuous datalog: window range leq 0}
A window of range $0$ over a stream $S$ at an instant 
${i \in \mathbb{T}}$ is always empty.
\begin{equation*}
W^{S,0}_{i} = \varnothing
\end{equation*}
\end{axiom}}

{\nobreak\begin{axiom}\label{axiom:continuous datalog: window start leq 0}
A window of any range $x \in \mathbb{N}$ over a stream $S$ at (or before)
the first instant of time is always empty.
\begin{equation*}
W^{S,x}_{i} = \varnothing
\end{equation*}
where $i \leqslant 0$.
\end{axiom}}

{\nobreak\begin{axiom}\label{axiom:continuous datalog: window composition}
A window of range ${x > 0}$ over a stream $S$ at an instant 
${i \in \mathbb{T}}$ is equal to the union of the $x$ consecutive
windows of range $1$ over the stream $S$ preceding and including that at
instant $i$.
\begin{equation*}
W^{S,x}_{i} = \mathop{\cup}_{j=0}^{x-1} W^{S,1}_{i-j}
\end{equation*}
\end{axiom}}

{\nobreak\begin{axiom}\label{axiom:continuous datalog: window disjointness}
A window of range ${x \in \mathbb{N}}$ over a stream $S$ at an
instant ${i \in \mathbb{T}}$ is disjoint from a second window of 
range ${y \in \mathbb{N}}$ over $S$ at any instant before $i - x$.
\begin{equation*}
W^{S,x}_{i} \cap W^{S,y}_{i-z} = \varnothing
\end{equation*}
where $z \geqslant x$.
\end{axiom}}

@@

We consider the entailment of the WC-Datalog program $\Pi^{C,r}$ not
as a single stream of fact instances (as for C-Datalog), but in three
parts: a set of fact instances $E^0 = cons^T(\langle R, F^0\rangle)$
that is the entailment of the TA-Datalog program $\langle R,
F^0\rangle$; a stream $P$ of fact instances entailed by the rules $R$ and
the fact instances in $F^0$ and $S$, ordered by the instant at which they
became justified (i.e. their instants); and a stream $N$ of those same fact
instances, ordered by the instant at which they ceased to be justified:
$\langle R, F^0, S, r \rangle \models \langle E^0, P, N \rangle$. $P$
and $N$ can be considered to be the continuous reasoning equivalents of
the streams produced by the \emph{IStream} and\emph{DStream} operators
respectively of the \emph{CQL} continuous query language \citep{CQL},
which create streams based on the additions to and deletions from a
relation. Note that $P$ is an in-order stream whose members are a subset
of those in $E^S$ for an equivalent C-Datalog program, while $N$ may be
an out-of-order stream.
\end{definition}

@@

{\nobreak\begin{axiom}\label{axiom:continuous datalog: entailment disjointness}
If two sliding windows over the same in-order stream $S$ cover
non-overlapping time periods, then the set of fact instances entailed
by ${\langle R, W^{S,x}_i \rangle}$ is necessarily disjoint from the
set of fact instances entailed by $\langle R, W^{S,y}_{i-z} \rangle$,
where ${z \geqslant x}$.
\[
\{ e \mid \langle R , W^{S,x}_i \rangle \models e \} \cap \{ e \mid \langle R , W^{S,y}_{i-z} \rangle \models e \} = \varnothing
\]
\end{axiom}}

@@

The instantaneous entailment of a $\Pi^{C,r}_i$ may be divided into two
semantically distinct portions: the \emph{persistent} portion is the
entailment of ${\langle R, F^0 \rangle \models E^0}$, as in a WC-Datalog
program, which is guaranteed to be part of the entailment of every state
due to the monotonicity of positive Datalog; the \emph{transitive} portion
is the set $E^T_{i}$ of entailed fact instances whose justifications are at
least partially contained in $W^{S,r}_i$. {\nobreak We define $E^T_i$ as:
\[
E^T_i = \{ x \mid \Pi^{C,r}_i \models x \land J^T \cup W^{S,r}_i \neq \varnothing \}
\]
where $J^T$ is a justification for the entailment of x by
$\Pi^{C,r}_i$.} By this definition and Axiom~\ref{axiom:continuous
  datalog: entailment disjointness}, $E^0$ is guaranteed to be
disjoint from every $E^T_i$ for every state $\Pi^{C,r}_i$ of every
WC-Datalog program $\Pi^{C,r}$.
\end{definition}

\begin{figure*}
\centering
$
\begin{CD}
  {\Pi^C = \langle R, F, S \rangle} @>\models>> {E^S} \\
  @V{add\ window}VV \\
        {\Pi^{C,r} = \langle R, F^0, S, r \rangle} @>\models>> {\langle cons^T(\langle R, F^0\rangle), P, N \rangle} \\
        @V{evaluate\ window}VV @VV{evaluate\ window}V \\
        {\Pi^{C,r}_i = \langle R, F^0 \cup W^{S,r}_i \rangle} @>\models>> {cons^T(\langle R, F^0 \rangle) \cup (W^{P,r}_i \setminus W^{N,r}_i)} \\
  @VVV @VVV \\
        {\Pi^T = \langle R, F^T \rangle } @>\models>> {cons^T(\Pi^T)} \\
  @V{strip}VV @VV{strip}V \\
        {\Pi = \langle R, F \rangle } @>\models>> {cons(\Pi)}
\end{CD}
$
\caption{Commutative diagram from Continuous Datalog programs to Datalog entailments.}
\labelfig{CD-TD-D}
\end{figure*}

@@

{\nobreak\begin{axiom}
\label{axiom:continuous datalog: positive window increment}
The window of range 1 at any given instant ${i \in \mathbb{T}}$ over
the stream $P$ entailed by a WC-Datalog program $\Pi^{C,r}$ is defined
as the set of fact instances that are entailed by the state
$\Pi^{C,r}_i$ that were not entailed by the state $\Pi^{C,r}_{i-1}$.
\begin{equation*}
W^{P,1}_{i} = \left( E^0 \cup E^T_{i} \right) \setminus
\left( E^0 \cup E^T_{i-1} \right) = E^T_{i} \setminus E^T_{i-1}
\end{equation*}
\end{axiom}}

{\nobreak\begin{axiom}
\label{axiom:continuous datalog: negative window increment}
The window of range 1 at any given instant ${i \in \mathbb{T}}$ over
the stream $N$ entailed by a WC-Datalog program $\Pi^{C,r}$ is defined
as the set of fact instances that were entailed by the state
$\Pi^{C,r}_{i-1}$ that are no longer entailed by the state
$\Pi^{C,r}_i$.
\begin{equation*}
W^{N,1}_{i} = \left( E^0 \cup E^T_{i-1} \right) \setminus
\left( E^0 \cup E^T_{i} \right) = E^T_{i-1} \setminus E^T_{i}
\end{equation*}
\end{axiom}}

{\nobreak\begin{axiom}
\label{axiom:continuous datalog: entailment precedes negation}
Fact instances may only be negated by some part of their justification
leaving the sliding window $W^{S,r}_{i}$, this necessarily being after
they were entailed.  As such, a window of any range ${x \in
  \mathbb{N}}$ over the stream of entailments $P$ at any given instant
${i \in \mathbb{T}}$ is disjoint from a window of any range ${y \in
  \mathbb{N}}$ over the stream of entailment negations $N$ at any
instant before ${i - x}$.
\begin{equation*}
  W^{P,x}_{i} \cap W^{N,y}_{i-z} = \varnothing
\end{equation*}
where $z \geqslant x$.
\end{axiom}}

@@

However, it follows from these axioms that the instantaneous set
$E^T_{i}$ of instances entailed at any given instant $i$ by any
$\Pi^{C,r} = \langle R, F^0, S, r \rangle$ may be reconstructed from
windows of sufficient range over its entailed streams $P$ and $N$ at
$i$:
\[ E^T_{i} = W^{P,x}_{i} \setminus W^{N,y}_{i} \]
where ${y \geqslant x \geqslant r}$. The minimum range $x$ is such to achieve
the \emph{complete} set of entailment instances $E^T_{i}$ of the
$\Pi^{C,r}_{i}$. In addition, the minimum range $y$ is such to achieve the
\emph{sound} set $E^T_{i}$. As such, a \emph{valid window} that
contains only and exactly those entailed instances $E^T_{i}$ at every
instant $i$ may be trivially maintained from the streams $P$ and $N$
continuously entailed by some $\Pi^{C,r}$.

@@

\end{nestedsection}
